Master's Project Journal

### 2019-03-25 conf_paper/paper.tex
tweaks and measurements

\title{FPGA Based Prototyping of an ADPLL Network.\\}

\author{\IEEEauthorblockN{C. Dooley, E. Blokhina, B. Mulkeen}
    \IEEEauthorblockA{\textit{School of Electrical \& Electronic Engineering, University College Dublin}
}

@@

@@

\section{System Architecture}
An ADPLL network, as the name would suggest, is created from a number of ADPLLs that are coupled using digital phase comparators which attempt to measure the phase error between two oscillators, with each non edge node being connected to four neighbours. Figure \ref{fig:adpll_network} illustrates such a system on a small scale, with a number of ADPLLs laid out in a Cartesian grid. The individual ADPLLs, often called distributed ADPLLs as the phase detectors are shared between nodes, are made up of a digitally controllable oscillator, an error combination block and a digital loop filter with the aforementioned digital phase detectors lying between each pair of oscillators.
\begin{figure}[h]
    \centering
    \includegraphics[width=0.25\textwidth]{adpll_network}
    \caption{ADPLL Network Architecture.}
    \label{fig:adpll_network}
    \vspace{0.5cm}
    \includegraphics[width=0.4\textwidth]{dist_adpll}
    \caption{Distributed ADPLL Design.}
    \label{fig:adpll_base}
\end{figure}
\subsection{Digitally Controlled Oscillators}
While regular Voltage Controlled Oscillator (VCO) is tuned by the input voltage over a continuous range of values, in a digital system there are a very limited number of voltages representable, most commonly just two levels. As such, using the voltage itself to control an oscillator is not a viable strategy and instead Digitally, or as they are sometimes known as Numerically, Controlled Oscillators (DCO) accept an \textit{n} bit wide control signal that alters the period of the output waveform. While tuning range, centre frequency and linearity apply as in a VCO, a DCO also has a minimum frequency step as a result of the non continuous control values. %TODO WORDING
In combination with the bit width of the control signal, the minimum frequency step limits the range over which the oscillator can be tuned. Frequently the output of the DCO is divided down to allow the control blocks in the feedback path that are clocked on this signal, to run at a lower and thus easier frequency.
\subsection{Loop Filter}
The loop filter is a key aspect of a phase lock loop as without it only the current output of the phase detector could be used to compute the control voltage of the VCO.
In an ADPLL it performs an identical role however, the digital environment results the traditional loop filter being replaced by a discrete time Proportional Integral (PI) controller, with the integral path of the controller implemented by an accumulator.
\subsection{Error Combiner}
The Error Combiner, as the name suggests, performs a summation of the error signals from the phase detectors lying between any neighbouring oscillator and its neighbours. It is often convenient for this to be performed as a weighted sum, thus avoiding different gain requirements for the loop filter of oscillators in the corners of the grid that only have a pair of neighbours. The ability to change the weights during operation enables ``uni-directional start-up'', which Javiden et. al \cite{javidan2011all} found prevented the system from entering unwanted stable equilibriums warned of in Pratt and Nguyen's seminal paper.
\subsection{Phase Detector}
As in a conventional PLL the phase detector outputs a value proportional to the difference in phase between a pair of signals, which is then output as a continuous voltage.
A digital phase detector then attempts to do the same by measuring the difference between fixed reference points in the waveform, which as a digital system deals with square waves, this is typically the rising edge of each input to the block.
Similarly to the oscillator, quantisation also plays a role in the behaviour of this block, with the time difference between each rising edge having a minimum possible measurement step.
In distributed ADPLLs the phase detector has dual outputs, each the inverse of the other, in order to feed the correct value to both of the connected error combiners.

@@

\subsubsection*{Design B}
Better suited for use with the inverter based oscillator, is a system that mimics the behaviour of a tapped delay line, working asynchronously to the FPGA clock. As the FPGA clock is not used, and that of the ADPLL cannot be used to measure itself, an unclocked sign detection circuit is required, an example of which can be seen in Figure \ref{fig:segn_pdetdl}. This is nearly identical to sign detection methods used on an ASIC, with the arbitration circuit replaced by a second SR Latch, which alleviates the issue of metastability.\\
A tapped delay line can be emulated on an FPGA through the use of inverter pairs, due to their short and approximately known delay placed between each tap.
The resolution then is the propagation time through the pair, although this will once again depend on the layout specifics, which was approximately 600 picoseconds on an Artix-7. This is significantly lower than the 4 nanoseconds achieved using the previous design and thus more suitable for use with an inverter ring oscillator, as this resolution is better than the time resolution of the oscillator. Figure \ref{fig:num_pdetdl} illustrates the configuration of the circuit used to emulate the function of a delay line that will be used in this paper.
\begin{figure}[h]
    \centering
    \includegraphics[width=0.4\textwidth]{../new_pdet1}
    \caption{Design B Sign Detection RTL Diagram.}
    \label{fig:segn_pdetdl}
\end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[width=0.4\textwidth]{../new_pdet2}
    \caption{Tapped Delay Line Emulation RTL Diagram.}
    \label{fig:num_pdetdl}
\end{figure}
\section{Comparison of Designs}
This paper will now examine the performance of a number of combinations of these blocks and the performance thereof, specifically oscillator 1 with phase detector A and oscillator 2 with detectors A \& B. For each test there were four ADPLLs on the FPGA arranged in a Cartesian Grid and as such each ADPLL will be referred to using matrix indexing corresponding to their grid location. For each configuration three different modes of operation were examined: Each ADPLL locking to the same 5 MHz external reference, each ADPLL connected to the left and above neighbours in a ``uni-directional'' network and finally a true ADPLL network where all neighbours are connected. For comparison, each configuration gains which had previously been determined to provide the best jitter performance were chosen. The data was captured using an Agilent MSO7054 A at 4 GSa/s. Each point of data is the mean of five 200 microsecond long captures which were then analysed in Matlab, Table \ref{table:b_c_perf} displays the results of this analysis.\\
The first columns of the table relate to the cycle-to-cycle jitter of the system, the difference between sequential rising edges, and the second portion gives the Time Interval Error, the difference between the actual timing of the edge and the ``ideal'' time as set by the reference. This combination of measurements will show how well each design performs the task of staying consistent from cycle to cycle both in relation to itself and the system at large.

\begin{table}[!ht]
    \begin{center}
        \begin{tiny}
            \begin{tabular}{cc|c|c|c|c|c|c|}           
                \cline{3-8}
                && \multicolumn{3}{c|}{Jitter Standard Deviation (ns)} & \multicolumn{3}{c|}{Time Interval Error. (ns)} \T\\
                \cline{3-8} 
                &&PLL 11&PLL 12&PLL 22&PLL 11&PLL 12&PLL 22\T\\
                \hline
                \multicolumn{2}{|l|}{Osc 1, PDET A}&-&-&-&-&-&-\T\\
                \multicolumn{2}{|r|}{Free PLLs}&1.5988&1.6178&1.6220&&& \T\\
                \multicolumn{2}{|r|}{Uni-dir.}&1.9804&2.0316&2.1317&&& \T\\
                \multicolumn{2}{|r|}{Bi-dir.}&2.0124&2.0681&2.0732&&& \T\\
                \hline
                \multicolumn{2}{|l|}{Osc 2, PDET A}&-&-&-&-&-&-\T\\
                \multicolumn{2}{|r|}{Free PLLs}&0.62499 $\mu$s&0.75919 $\mu$s&0.79049  $\mu$s&&& \T\\
                \multicolumn{2}{|r|}{Uni-dir.}&0.76822 $\mu$s&0.9009 $\mu$s&1.1407 $\mu$s&&& \T\\
                \multicolumn{2}{|r|}{Bi-dir.}&0.70393 $\mu$s&0.70201 $\mu$s&1.241 $\mu$s&&& \T\\
                \hline
                \multicolumn{2}{|l|}{Osc 2, PDET B}&-&-&-&-&-&-\T\\
                \multicolumn{2}{|r|}{Free PLLs}&4.1680&3.9803&5.0741&&& \T\\
                \multicolumn{2}{|r|}{Uni-dir.}&4.3820&5.7315&6.2336 &&& \T\\
                \multicolumn{2}{|r|}{Bi-dir.}&5.6137&7.1831&7.1612 &&& \T\\
                \hline
                \B
            \end{tabular}
        \end{tiny}
        \caption{}
        \label{table:b_c_perf}
    \end{center}
\end{table}

From the table it is apparent that the combination of the inverter chain DCO and the FPGA clock driven phase detector is an ill-advised choice, as the jitter present in the system is several orders of magnitude worse than either other situation. In this system, the FPGA clock is at 258 MHz, the maximum possible, thus giving a detection resolution of 3.875 nanoseconds. This is 3.5 times greater than the period resolution of approximately 1.2 nano second period resolution of the oscillator, and requires the gains of the PI filter to be increased fourfold in order to respond appropriately to the error detected. The problems of relatively inaccurate detection and increased gains combine to produce a system that is only able to achieve poor locking performance.
Both other combinations work well however, delivering a level of jitter that is both low yet allows an indication of how impactful variations in architecture or control parameters are to be seen visually on an oscilloscope.\\

\begin{figure}[h]%
    \centering
    \subfloat[Inverter Based Design.]{{\includegraphics[width=0.5\linewidth]{distrib_ring} }}%
    \subfloat[FPGA Clocked Design.]{{\includegraphics[width=0.5\linewidth]{distrib_pa} }}%
    \caption{Example Distribution of Periods.\label{fig:dists}}    
\end{figure}
%\begin{figure}[h]
%   \centering
%   \subfloat[FPGA Clocked Design][width=0.25\textwidth]{distrib_pa}
%   \caption{Distribution of Periods.}
%   \label{fig:dist2}
%\end{figure}
Figure \ref{fig:dists} paints a picture of the unrealistic nature of the FPGA clocked designs as in comparison to the ring based design, each oscillator will have identical intrinsic period steps thus eliminating any jitter that maybe be seen in the system due to misalignment. This effects the suitability of the FPGA clocked designs for the improvement of simulation models, or the examination of how a novel block may impact the behaviour of the network.
\section*{Conclusion}
An FPGA is a suitable environment for the prototyping of ADPLL networks and can fill a number of roles depending on the designer's requirements. However, the designer should bear in mind the platform's limitations, especially when attempting to use a frequency that is not orders of magnitude lower than the maximum clock available from the FPGA's clock manager. Inverter based and FPGA clocked designs, both have a niche in which they are the preferred option, with the clocked design better suiting replication of an existing design at scaled-down frequencies and the inverter based designs better in the modelling and analysis of ADPLL network behaviour.

>>>>

\title{FPGA Based Prototyping of an ADPLL Network\\}

\author{\IEEEauthorblockN{C. Dooley, E. Blokhina, B. Mulkeen}
    \IEEEauthorblockA{\textit{School of Electrical \& Electronic Engineering, University College Dublin}
	}
}

@@

\vspace{-4.5cm}

@@

\section{System Architecture}
An ADPLL network, as the name would suggest, is created from a number of ADPLLs that are coupled using digital phase comparators which attempt to measure the phase error between two oscillators, with each non edge node being connected to four neighbours. Figure \ref{fig:adpll_network} illustrates such a system on a small scale, with a number of ADPLLs laid out in a Cartesian grid. The individual ADPLLs, often called distributed ADPLLs as the phase detectors are shared between nodes, are made up of a digitally controllable oscillator, an error combination block and a digital loop filter with the aforementioned digital phase detectors lying between each pair of oscillators.
\begin{figure}[h]
    \centering
    \includegraphics[width=0.25\textwidth]{adpll_network}
    \caption{ADPLL Network Architecture.}
    \label{fig:adpll_network}
    \vspace{0.5cm}
    \includegraphics[width=0.4\textwidth]{dist_adpll}
    \caption{Distributed ADPLL Design.}
    \label{fig:adpll_base}
    \vspace{-0.5cm}
\end{figure}
\subsection{Digitally Controlled Oscillators}
While regular Voltage Controlled Oscillator (VCO) is tuned by the input voltage over a continuous range of values, in a digital system there are a very limited number of voltages representable, most commonly just two levels. As such, using the voltage itself to control an oscillator is not a viable strategy and instead Digitally, or as they are sometimes known as Numerically, Controlled Oscillators (DCO) accept an \textit{n} bit wide control signal that alters the period of the output waveform. While tuning range, centre frequency and linearity apply as in a VCO, a DCO also has a minimum frequency step as a result of the non continuous control values. %TODO WORDING
In combination with the bit width of the control signal, the minimum frequency step limits the range over which the oscillator can be tuned. Frequently the output of the DCO is divided down to allow the control blocks in the feedback path that are clocked on this signal, to run at a lower and thus easier frequency.
\subsection{Loop Filter}
The loop filter is a key aspect of a phase lock loop as without it only the current output of the phase detector could be used to compute the control voltage of the VCO.
In an ADPLL it performs an identical role however, the digital environment results the traditional loop filter being replaced by a discrete time Proportional Integral (PI) controller, with the integral path of the controller implemented by an accumulator.
\subsection{Error Combiner}
The Error Combiner, as the name suggests, performs a summation of the error signals from the phase detectors lying between any neighbouring oscillator and its neighbours. It is often convenient for this to be performed as a weighted sum, thus avoiding different gain requirements for the loop filter of oscillators in the corners of the grid that only have a pair of neighbours. The ability to change the weights during operation enables ``uni-directional start-up'', which Javiden et. al \cite{javidan2011all} found prevented the system from entering unwanted stable equilibriums warned of in Pratt and Nguyen's seminal paper.
\subsection{Phase Detector}
As in a conventional PLL the phase detector outputs a value proportional to the difference in phase between a pair of signals, which is then output as a continuous voltage.
A digital phase detector then attempts to do the same by measuring the difference between fixed reference points in the waveform, which as a digital system deals with square waves, this is typically the rising edge of each input to the block.
Similarly to the oscillator, quantisation also plays a role in the behaviour of this block, with the time difference between each rising edge having a minimum possible measurement step.
In distributed ADPLLs the phase detector has dual outputs, each the inverse of the other, in order to feed the correct value to both of the connected error combiners.

@@

\subsubsection*{Design B}
Better suited for use with the inverter based oscillator, is a system that mimics the behaviour of a tapped delay line, working asynchronously to the FPGA clock. As the FPGA clock is not used, and that of the ADPLL cannot be used to measure itself, an unclocked sign detection circuit is required, an example of which can be seen in Figure \ref{fig:segn_pdetdl}. This is nearly identical to sign detection methods used on an ASIC, with the arbitration circuit replaced by a second SR Latch, which alleviates the issue of metastability.\\
A tapped delay line can be emulated on an FPGA through the use of inverter pairs, due to their short and approximately known delay placed between each tap.
The resolution then is the propagation time through the pair, although this will once again depend on the layout specifics, which was approximately 600 picoseconds on an Artix-7. This is significantly lower than the 4 nanoseconds achieved using the previous design and thus more suitable for use with an inverter ring oscillator, as this resolution is better than the time resolution of the oscillator. Figure \ref{fig:num_pdetdl} illustrates the configuration of the circuit used to emulate the function of a delay line that will be used in this paper.
\begin{figure}[h]
    \centering
    \includegraphics[width=0.4\textwidth]{../new_pdet1}
    \caption{Design B Sign Detection RTL Diagram.}
    \label{fig:segn_pdetdl}
    \vspace{-0.5cm}
\end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[width=0.4\textwidth]{../new_pdet2}
    \caption{Tapped Delay Line Emulation RTL Diagram.}
    \label{fig:num_pdetdl}
    \vspace{-0.5cm}
\end{figure}
\section{Comparison of Designs}
This paper will now examine the performance of a number of combinations of these blocks and the performance thereof, specifically oscillator 1 with phase detector A and oscillator 2 with detectors A \& B. For each test there were four ADPLLs on the FPGA arranged in a Cartesian Grid and as such each ADPLL will be referred to using matrix indexing corresponding to their grid location. For each configuration three different modes of operation were examined: Each ADPLL locking to the same 5 MHz external reference, each ADPLL connected to the left and above neighbours in a ``uni-directional'' network and finally a true ADPLL network where all neighbours are connected. For comparison, each configuration gains which had previously been determined to provide the best jitter performance were chosen. The data was captured using an Agilent MSO7054 A at 4 GSa/s. Each point of data is the mean of five 200 microsecond long captures which were then analysed in Matlab, Table \ref{table:b_c_perf} displays the results of this analysis.\\
The first columns of the table relate to the cycle-to-cycle jitter of the system, the difference between sequential rising edges, and the second portion gives the Time Interval Error, the difference between the actual timing of the edge and the ``ideal'' time as set by the reference. This combination of measurements will show how well each design performs the task of staying consistent from cycle to cycle both in relation to itself and the system at large.

\begin{table}[!ht]
    \begin{center}
        \begin{tiny}
        	\setlength{\tabcolsep}{.9\tabcolsep}
            \begin{tabular}{cc|c|c|c|c|c|c|}           
                \cline{3-8}
                && \multicolumn{3}{c|}{Jitter Standard Deviation (ns)} & \multicolumn{3}{c|}{Max. Time Interval Error. (ns)} \T\\
                \cline{3-8} 
                &&PLL 11&PLL 12&PLL 22&PLL 11&PLL 12&PLL 22\T\\
                \hline
                \multicolumn{2}{|l|}{Osc 1, PDET A}&-&-&-&-&-&-\T\\
                \multicolumn{2}{|r|}{Free PLLs} &1.4377 &1.4714&1.4217   &6.1910&6.3667&6.8052 \T\\
                \multicolumn{2}{|r|}{Uni-dir.}  &0.91589&1.0993&1.0875   &8.3619&8.9768&10.974 \T\\
                \multicolumn{2}{|r|}{Bi-dir.}   &1.0470 &1.0979&1.0306   &24.998&33.767&33.204 \T\\
                \hline
                \multicolumn{2}{|l|}{Osc 2, PDET A}&-&-&-&-&-&-\T\\
                \multicolumn{2}{|r|}{Free PLLs}&&& 0&&& \T\\
                \multicolumn{2}{|r|}{Uni-dir.}&&& 0&&& \T\\
                \multicolumn{2}{|r|}{Bi-dir.}&&& 0&&& \T\\
                \hline
                \multicolumn{2}{|l|}{Osc 2, PDET B}&-&-&-&-&-&-\T\\
                \multicolumn{2}{|r|}{Free PLLs} &7.8090&7.8827&6.6272   &17.336&21.151&24.302 \T\\
                \multicolumn{2}{|r|}{Uni-dir.}  &7.7866&8.1145&7.3598   &17.649&29.962&37.051 \T\\
                \multicolumn{2}{|r|}{Bi-dir.}   &7.7575&7.7729&6.6962   &30.019&40.577&45.617 \T\\
                \hline
                \B
            \end{tabular}
        \end{tiny}
        \caption{}
        \label{table:b_c_perf}
    \end{center}
\vspace{-0.5cm}
\end{table}
From the table it is apparent that the combination of the inverter chain DCO and the FPGA clock driven phase detector is an ill-advised choice, as the jitter present in the system is several orders of magnitude worse than either other situation. In this system, the FPGA clock is at 258 MHz, the maximum possible, thus giving a detection resolution of 3.875 nanoseconds. This is 3.5 times greater than the period resolution of approximately 1.2 nano second period resolution of the oscillator, and requires the gains of the PI filter to be increased fourfold in order to respond appropriately to the error detected. The problems of relatively inaccurate detection and increased gains combine to produce a system that is only able to achieve poor locking performance.
Both other combinations work well however, delivering a level of jitter that is both low yet allows an indication of how impactful variations in architecture or control parameters are to be seen visually on an oscilloscope.\\

\begin{figure}[h]%
    \centering
    \subfloat[Inverter Based Design.]{{\includegraphics[width=0.5\linewidth]{distrib_ring} }}%
    \subfloat[FPGA Clocked Design.]{{\includegraphics[width=0.5\linewidth]{distrib_pa} }}%
    \caption{Example Distribution of Periods.\label{fig:dists}}    
    \vspace{-0.5cm}
\end{figure}
%\begin{figure}[h]
%   \centering
%   \subfloat[FPGA Clocked Design][width=0.25\textwidth]{distrib_pa}
%   \caption{Distribution of Periods.}
%   \label{fig:dist2}
%\end{figure}
Figure \ref{fig:dists} paints a picture of the unrealistic nature of the FPGA clocked designs as in comparison to the ring based design, each oscillator will have identical intrinsic period steps thus eliminating any jitter that maybe be seen in the system due to misalignment. This effects the suitability of the FPGA clocked designs for the improvement of simulation models, or the examination of how a novel block may impact the behaviour of the network.
\section*{Conclusion}
An FPGA is a suitable environment for the prototyping of ADPLL networks and can fill a number of roles depending on the designer's requirements. However, the designer should bear in mind the platform's limitations, especially when attempting to use a frequency that is not orders of magnitude lower than the maximum clock available from the FPGA's clock manager. Inverter based and FPGA clocked designs, both have a niche in which they are the preferred option, with the clocked design better suiting replication of an existing design at scaled-down frequencies and the inverter based designs better in the modelling and analysis of ADPLL network behaviour.
